Approximation as Differentiation

### 2025-05-08 main/formal-structures.tex
Recap functor categories [X, C] where X a set.

For a total language where types denote sets, the goal is to interpret programs as Galois connections relating
approximations of their input to approximations of their output. Since these lattice of approximations depends
on the value being approximated, the semantics of every type $\tau$ must have both a ``conventional''
component, assigning $\tau$ to a set $X$, and an associated lattice of approximations $\partial X(x)$ for
every point $x \in X$. The semantics of expression $e$ of type $Y$ with a single free variable of type $X$
must then also have two components: a function $f: X \to Y$ giving the baseline (unapproximated) meaning of
the expression, plus a family of Galois connections $\partial f(x): \partial X(x) \to \partial Y(f(x))$ for
every $x \in X$ which can be used for slicing over that expression.

This pattern of objects and morphisms is exactly the \emph{Grothendieck construction} for a particular indexed
category $T: \Set^\op \to \Cat$, namely the functor which assigns to every set $X$ the functor category $T(X)
= \Func{X}{\LatGal}$. Here the (discrete) category $X$ is a plain set and so $\Func{X}{\LatGal}$ is exactly
the category of $X$-indexed families of bounded lattices, with $X$-indexed families of Galois connections
between them as morphisms. The Grothendieck construction $\Grothendieck{}T$ is the ``total category'' which
gloms all such $\Set$-indexed families together, with reindexing morphisms to transport between different
indexing sets.

@@

As sketched in \secref{introduction:galois-slicing}, Galois slicing requires that the approximants of a point
$x$ form a bounded lattice $(X, \meet, \join, \top, \bot)$, with least element $\bot$ representing the
approximation that discards all information about $x$, greatest element $\top$ representing the approximation
that retains all information about $x$, and $\meet$ and $\join$ providing two canonical ways to combine
approximations. Thus rather than working directly in $\PosGal$, we define the following subcategory.

\begin{definition}
Define the category $\LatGal$ which has as objects $X = (X, \meet, \join, \top, \bot)$ all bounded lattices,
and as morphisms all Galois connections $f \adj g: X \to Y$.
\end{definition}

@@

>>>>

@@

As sketched in \secref{introduction:galois-slicing}, Galois slicing requires that the approximants of a point
$x$ form a bounded lattice $(X, \meet, \join, \top, \bot)$, with least element $\bot$ representing the
approximation that discards all information about $x$, greatest element $\top$ representing the approximation
that retains all information about $x$, and $\meet$ and $\join$ providing two canonical ways to combine
approximations. Thus rather than working directly in $\PosGal$, we consider the following subcategory instead.

\begin{definition}
Define $\LatGal$ to be the category which has as objects $X = (X, \meet, \join, \top, \bot)$ all bounded
lattices, and as morphisms all Galois connections $f \adj g: X \to Y$.
\end{definition}

@@

For a total language where types denote sets, our goal is to interpret programs as Galois connections relating
approximations of their inputs to approximations of their outputs. Since these lattice of approximations
depends on the value being approximated, the semantics of every type $\tau$ must have both a ``conventional''
component, assigning $\tau$ to a set $X$, and an associated lattice of approximations $\partial X(x)$ for
every point $x \in X$. The semantics of expression $e$ of type $Y$ with a single free variable of type $X$
must then also have two components: a function $f: X \to Y$ giving the baseline (unapproximated) meaning of
the expression, plus a family of Galois connections $\partial f(x): \partial X(x) \to \partial Y(f(x))$ for
every $x \in X$ which can be used for slicing over that expression.

This pattern of objects and morphisms is exactly the \emph{Grothendieck construction} for a particular indexed
category $T: \Set^\op \to \Cat$, namely the functor which assigns to every set $X$ the functor category $T(X)
= \Func{X}{\LatGal}$. Here the (discrete) category $X$ is a plain set and so $\Func{X}{\LatGal}$ is exactly
the category of $X$-indexed families of bounded lattices, with $X$-indexed families of Galois connections
between them as morphisms. The Grothendieck construction $\Grothendieck{}T$ is the ``total category'' which
gloms all such $\Set$-indexed families of bounded lattices into the same category, with reindexing morphisms
to transport between different indexing sets.

\begin{definition}[Category of $X$-indexed families of objects]
\label{def:fam:Fam}
For any set $X$ (interpreted as a discrete category) and any category $\cat{C}$, recall that the functor
category $\Func{X}{C}$ has as objects all functors $X \to \cat{C}$, in other words all $X$-indexed families of
objects of $\cat{C}$, and as morphisms from $X \to \cat{C}$ to $Y \to \cat{C}$, all natural transformations,
i.e.~families of morphisms $f_x: X_x \to Y_x$ in $\cat{C}$ for every $x \in X$.
\end{definition}
